---
title: "Practical Machine Learning Project"
author: Tae Seung Kang
date: "October 25, 2015"
output:
  html_document:
    fig_height: 9
    fig_width: 9
  pdf_document:
    fig_height: 9
    fig_width: 9
---

## Overview
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, our goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

The goal of this project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. In this report, we describe how we built our model, how we used cross validation, what the expected out of sample error is, and why we made the choices we did. 


## Data Preparation

The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. 

```{r, cache = T, echo=F}
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(corrplot)

os <- .Platform$OS.type
trainUrl <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv" # for linux
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv" # for linux
if(os == "windows") {
  trainUrl <-"http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv" # for windows
  testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv" # for windows
} 
trainFile <- "./data/pml-training.csv"
testFile  <- "./data/pml-testing.csv"
if (!file.exists("./data")) {
  dir.create("./data")
}
if (!file.exists(trainFile)) {
  if(os == "windows") 
    download.file(trainUrl, destfile=trainFile, method = "internal") # for windows
  else
    download.file(trainUrl, destfile=trainFile, method="curl") # for linux
}
if (!file.exists(testFile)) {
  if(os == "windows") 
    download.file(testUrl, destfile=testFile, method = "internal") # for windows
  else
    download.file(testUrl, destfile=testFile, method="curl") # for linux
}
```  

After downloading, we load the training and testing data sets assuming they are in directory "./data".
```{r, cache = T}
trainset <- read.csv(trainFile)
testset <- read.csv(testFile)
dim(trainset)
dim(testset)
```
The training data set contains 19622 observations and 160 variables, while the testing data set contains 20 observations and 160 variables. The "classe" variable in the testing set is the outcome to predict.

### Data Cleaning
In this step, we clean the data and get rid of missing values as well as some meaningless variables.
```{r, cache = T, echo=F, results="hide"}
sum(complete.cases(trainset))
```
First, we remove the columns that contain missing values NA.
```{r, cache = T}
trainset <- trainset[, colSums(is.na(trainset)) == 0] 
testset <- testset[, colSums(is.na(testset)) == 0] 
```  
Next, we get rid of some columns that do not contribute much to the accelerometer measurements.
```{r, cache = T}
classe <- trainset$classe
trainRemove <- grepl("^X|timestamp|window", names(trainset))
trainset <- trainset[, !trainRemove]
trainCleaned <- trainset[, sapply(trainset, is.numeric)]
trainCleaned$classe <- classe
testRemove <- grepl("^X|timestamp|window", names(testset))
testset <- testset[, !testRemove]
testCleaned <- testset[, sapply(testset, is.numeric)]
```
Now, the cleaned training data set contains 19622 observations with 53 variables and the testing data set contains 20 observations with 53 variables. The "classe" variable is still in the cleaned training set.

### Data Splitting
Then, we can split the cleaned training set into a pure training data set (70%) and a validation data set (30%). We will use the validation data set to conduct cross validation in future steps.  
```{r, cache = T}
# Set a seed
set.seed(500) 
inTrain <- createDataPartition(trainCleaned$classe, p=0.70, list=F)
trainset <- trainCleaned[inTrain, ]
testset <- trainCleaned[-inTrain, ]
```

## Data Modeling
We fit a predictive model for activity recognition using **Random Forest** algorithm because it automatically selects important variables and is robust to correlated covariates & outliers in general. We will use **5-fold cross validation** when applying the algorithm.  
```{r, cache = T}
controlRf <- trainControl(method="cv", 5)
modelRf <- train(classe ~ ., data=trainset, method="rf", trControl=controlRf, ntree=250)
modelRf
```
Then, we estimate the performance of the model on the validation data set.  
```{r, cache = T}
predictRf <- predict(modelRf, testset)
confusionMatrix(testset$classe, predictRf)
```
```{r, cache = T}
accuracy <- postResample(predictRf, testset$classe)
accuracy
oose <- 1 - as.numeric(confusionMatrix(testset$classe, predictRf)$overall[1])
oose
```
So, the estimated accuracy of the model is 99.42% and the estimated out-of-sample error is 0.58%.

## Predicting for Test Data Set
Now, we apply the model to the original testing data set downloaded from the data source. We remove the `problem_id` column first.  
```{r, cache = T}
result <- predict(modelRf, testCleaned[, -length(names(testCleaned))])
result
```  

## Appendix: Figures
1. Correlation Matrix Visualization  
```{r, cache = T}
corrPlot <- cor(trainset[, -length(names(trainset))])
corrplot(corrPlot, method="color")
```
2. Decision Tree Visualization
```{r, cache = T}
treeModel <- rpart(classe ~ ., data=trainset, method="class")
prp(treeModel) # fast plot
```


